---
title: "Estimation of Market Impact"
format:
  pdf:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 4.5
author: "Raj Kundi"
date: "08/21/2026"
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(ggplot2)
library(tidyverse)
library(data.table)
library(lubridate)
library(zoo)
library(scales)

setDTthreads(0)

# For PDF output, use cairo_pdf rather than the default pdf() device: pdf()
# cannot encode non-ASCII characters (Greek letters, em-dashes) in plot labels
# and drops them silently with only a warning. cairo_pdf handles UTF-8 and is
# still vector output. Guarded so HTML rendering is unaffected.
if (knitr::is_latex_output()) knitr::opts_chunk$set(dev = "cairo_pdf")

# Format a number as "$m \times 10^{e}$" for correctly rendered scientific notation inline.
format_sci <- function(x, digits = 3) {
  exponent <- floor(log10(abs(x)))
  mantissa <- signif(x / 10^exponent, digits)
  sprintf("$%s \\times 10^{%d}$", format(mantissa), exponent)
}

# ---- Plot styling -----------------------------------------------------------
# Categorical hues are slots 1-3 of a colourblind-safe reference palette, used in
# fixed order: blue = measured data / linear fit, orange = power law, grey = context.
col_data  <- "#2a78d6"
col_model <- "#eb6834"
col_muted <- "#b4b3ac"

theme_set(theme_minimal(base_size = 11))
theme_update(
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(linewidth = 0.3, colour = "grey88"),
  plot.title       = element_text(face = "bold", size = 12),
  plot.subtitle    = element_text(colour = "grey35", size = 10),
  plot.caption     = element_text(colour = "grey45", size = 8, hjust = 0),
  strip.text       = element_text(colour = "grey25", size = 9),
  legend.position  = "bottom",
  legend.title     = element_blank()
)
```

```{r}
#| label: parameters
#| code-summary: "Estimation parameters"

SYMBOL       <- "SOLUSD_PERP"
MONTH        <- "2024-08"

BAR_MINUTES  <- 5      # time-bar length
VOL_BAR_SIZE <- 5000   # volume-bar size, in base_qty (SOL)
VOL_WINDOW   <- 15     # bars in the rolling volatility estimate used for filtering
TRIM_Q       <- 0.95   # quantile at which volatility, volume and imbalance are trimmed
N_BINS       <- 20     # imbalance quantile bins for the conditional mean
HAC_LAG      <- 12     # Newey-West lag (1 hour at 5-minute bars)
```

# Introduction

To compute trading curves and trading envelopes, market impact and execution
costs are needed. This notebook estimates the **permanent** impact parameter
$k$; the companion notebook `../execution_cost/execution_cost_function.qmd`
estimates the *temporary* impact parameters from the same data.

The mid-price of the asset is modelled by the process

$$
d S_t = \sigma \, d W_t + k\, v_t \, dt,
$$

where $\sigma$ is the arithmetic volatility and $k \ge 0$ is the parameter
governing **permanent market impact**. Integrating the drift over a bar gives

$$
S_{t+\Delta} - S_t \approx k \sum_i v_i + \text{noise},
$$

so regressing a bar's mid-price change (`price_diff`) on its net signed volume
(`imbalance`) recovers $k$ as the slope. The regression is run without an
intercept, since the model carries no drift term.

The distinction between the two impact parameters is what each one is measured
against, and it matters for how the estimate is built:

| | Permanent impact ($k$, here) | Temporary impact (companion notebook) |
|---|---|---|
| What moves | the mid-price itself | the execution price relative to the touch |
| Measured over | a bar of many trades | a single marketable order |
| Persists? | yes, by definition — and this is tested below | no, it decays as the book refills |

Two things are worth stating up front, because they shape the notebook. First,
permanence is an *assumption* in the model above, and an assumption that can be
checked: [Does the impact actually persist?] does so directly. Second, the model
asserts that impact is *linear* in order flow, and
[Is impact linear in imbalance?] shows that it is not — which turns out to
explain why the estimate of $k$ is sensitive to the cleaning choices.

# Data

We use one month (`r MONTH`) of Binance `r SYMBOL` perpetual futures data:
tick-level `trades` (individual executions, signed by taker side) and
`bookTicker` (best bid/ask snapshots), from which the mid-price process is
reconstructed. Only the three `bookTicker` columns needed are read, since the
raw file is several gigabytes.

```{r}
#| label: load-data
#| message: false
#| code-summary: "Load trades and bookTicker"

trades <- fread(sprintf("data/%s-trades-%s.csv", SYMBOL, MONTH))
bookTicker <- fread(
  sprintf("data/%s-bookTicker-%s.csv", SYMBOL, MONTH),
  select = c("best_bid_price", "best_ask_price", "transaction_time")
)

trades[, ts := as.POSIXct(time / 1000, origin = "1970-01-01", tz = "UTC")]
bookTicker[, mid_price := (best_bid_price + best_ask_price) / 2]

setorder(trades, time, id)
setorder(bookTicker, transaction_time)
```

For each trade we attach the most recent mid-price observed before it
(`pre_mid`) and the first mid-price observed after it (`post_mid`), so that the
difference between the two isolates the price move associated with that trade
(or a bar of trades) net of unrelated mid-price drift.

```{r}
#| label: quote-join
#| code-summary: "Non-equi join of trades to the surrounding mid-prices"

setkey(bookTicker, transaction_time)

trades[, pre_mid  := bookTicker[trades, on = .(transaction_time < time),
                                mult = "last",  .(mid_price)]$mid_price]
trades[, post_mid := bookTicker[trades, on = .(transaction_time > time),
                                mult = "first", .(mid_price)]$mid_price]
```

```{r}
#| label: free-bookticker
#| include: false

rm(bookTicker)
invisible(gc())
```

The taker side signs the flow: `is_buyer_maker == TRUE` means the maker was the
buyer, so the taker was *selling*.

```{r}
#| label: sign-flow
#| code-summary: "Sign the order flow by taker side"

trades[, side_sign  := fifelse(is_buyer_maker, -1, 1)]
trades[, signed_qty := base_qty * side_sign]
trades[, cum_base_qty := cumsum(base_qty)]

mean_price <- trades[, sum(base_qty * price) / sum(base_qty)]
```

# Bar construction

Bars are constructed two ways, to check that the estimate of $k$ is not an
artifact of the sampling scheme: fixed `r BAR_MINUTES`-minute **time bars**, and
**volume bars** that close every `r comma(VOL_BAR_SIZE)` units of `base_qty`
traded. Both use the same aggregation, so it is written once and called twice.

```{r}
#| label: make-bars
#| code-summary: "Bar aggregation"

make_bars <- function(bin_id) {
  b <- trades[, .(
    imbalance      = sum(signed_qty),
    base_qty       = sum(base_qty),
    n_trades       = .N,
    pre_mid_price  = first(pre_mid),
    post_mid_price = last(post_mid)
  ), by = .(bin = bin_id)]

  # Explicit ordering: the rolling volatility below assumes chronological order,
  # and relying on the grouping to deliver it is an implicit dependency.
  setorder(b, bin)
  b[, price_diff := post_mid_price - pre_mid_price]
  b[, std_dev    := rollapply(pre_mid_price, VOL_WINDOW, sd, fill = NA, align = "right")]
  b[]
}

bars_time_all   <- make_bars(floor_date(trades$ts, sprintf("%d minutes", BAR_MINUTES)))
bars_volume_all <- make_bars(floor(trades$cum_base_qty / VOL_BAR_SIZE))
```

The `r BAR_MINUTES`-minute grid turns out to be complete — every one of the
`r comma(nrow(bars_time_all))` bins in the month contains at least one trade,
which matters for the decay test below, since that shifts the bar series
forward in time and would otherwise skip over gaps.

```{r}
#| label: contiguity
#| include: false

bins_contiguous <- all(diff(as.numeric(bars_time_all$bin)) == BAR_MINUTES * 60)
stopifnot(bins_contiguous)
```

# Data cleaning

Three filters are applied. The top `r sprintf("%.0f%%", 100 * (1 - TRIM_Q))` of
rolling volatility (`std_dev`) and the top
`r sprintf("%.0f%%", 100 * (1 - TRIM_Q))` of traded volume (`base_qty`) are
dropped, to exclude bars dominated by news or liquidation events. The top and
bottom `r sprintf("%.0f%%", 100 * (1 - TRIM_Q))` of `imbalance` are dropped to
limit outlier influence on the OLS fit.

```{r}
#| label: filters
#| code-summary: "Filtering"

apply_filters <- function(b, q = TRIM_Q) {
  cut_sd  <- quantile(b$std_dev,   q,     na.rm = TRUE)
  cut_qty <- quantile(b$base_qty,  q,     na.rm = TRUE)
  cut_hi  <- quantile(b$imbalance, q,     na.rm = TRUE)
  cut_lo  <- quantile(b$imbalance, 1 - q, na.rm = TRUE)
  b[!is.na(std_dev) & std_dev <= cut_sd & base_qty <= cut_qty &
      imbalance <= cut_hi & imbalance >= cut_lo]
}

# Note the separate names: apply_filters is deliberately not applied in place.
# Overwriting the source table would make re-running this chunk re-trim an
# already-trimmed sample, with the quantiles recomputed on the smaller set.
bars_time   <- apply_filters(bars_time_all)
bars_volume <- apply_filters(bars_volume_all)
```

::: {.callout-note}
## A filter that was doing nothing

An earlier version of this notebook also dropped bars with
`abs(imbalance) <= 1`, on the grounds that they carry little order-flow
information. On this sample that removed 12 bars out of
`r comma(nrow(bars_time))` and changed $k$ in the fourth significant figure, so
it has been removed. (The threshold was also described as "roughly $100
notional"; 1 SOL was about `r sprintf("$%.0f", mean_price)` over this period.)
:::

```{r}
#| label: cleaning-table
#| code-summary: "Cleaning waterfall"

waterfall <- function(b, label) {
  cut_sd  <- quantile(b$std_dev,   TRIM_Q,     na.rm = TRUE)
  cut_qty <- quantile(b$base_qty,  TRIM_Q,     na.rm = TRUE)
  cut_hi  <- quantile(b$imbalance, TRIM_Q,     na.rm = TRUE)
  cut_lo  <- quantile(b$imbalance, 1 - TRIM_Q, na.rm = TRUE)
  d_sd  <- b[is.na(std_dev) | std_dev > cut_sd, .N]
  keep1 <- b[!is.na(std_dev) & std_dev <= cut_sd]
  d_qty <- keep1[base_qty > cut_qty, .N]
  keep2 <- keep1[base_qty <= cut_qty]
  d_imb <- keep2[imbalance > cut_hi | imbalance < cut_lo, .N]
  data.table(Sample = label, Constructed = nrow(b),
             `Volatility` = -d_sd, `Volume` = -d_qty, `Imbalance tails` = -d_imb,
             Retained = nrow(b) - d_sd - d_qty - d_imb)
}

knitr::kable(
  rbind(waterfall(bars_time_all,   sprintf("%d-minute bars", BAR_MINUTES)),
        waterfall(bars_volume_all, sprintf("%s-SOL volume bars", comma(VOL_BAR_SIZE)))),
  format.args = list(big.mark = ","),
  caption = "Bars removed by each cleaning step."
)
```

# Regression

The slope on `imbalance` estimates $k$. Because bars overlap in information and
order flow is autocorrelated, OLS standard errors understate the true
uncertainty, so **Newey--West (HAC) standard errors** are reported alongside
them, at a lag of `r HAC_LAG` bars.

```{r}
#| label: hac
#| code-summary: "Newey-West standard errors"

# Hand-rolled rather than via sandwich::NeweyWest, to avoid a dependency for
# what is a dozen lines: Bartlett-weighted sum of autocovariances of the score.
nw_se <- function(m, L = HAC_LAG) {
  X <- model.matrix(m)
  u <- as.vector(residuals(m))
  n <- nrow(X)
  bread <- solve(crossprod(X))
  meat  <- crossprod(X * u)
  for (l in seq_len(L)) {
    w <- 1 - l / (L + 1)
    G <- crossprod(X[(l + 1):n, , drop = FALSE] * u[(l + 1):n],
                   X[1:(n - l), , drop = FALSE] * u[1:(n - l)])
    meat <- meat + w * (G + t(G))
  }
  sqrt(diag(bread %*% meat %*% bread))
}
```

## Time bars (`r BAR_MINUTES` minutes)

```{r}
#| label: fit-time
#| code-summary: "Fit on time bars"

fit_time <- lm(price_diff ~ imbalance + 0, bars_time)
summary(fit_time)
```

## Volume bars (`r comma(VOL_BAR_SIZE)` base_qty)

```{r}
#| label: fit-volume
#| code-summary: "Fit on volume bars"

fit_volume <- lm(price_diff ~ imbalance + 0, bars_volume)
summary(fit_volume)
```

```{r}
#| label: fit-table
#| code-summary: "Both specifications with HAC errors"

fit_summary <- rbindlist(list(
  data.table(Bars = sprintf("%d-minute time bars", BAR_MINUTES), n = nrow(bars_time),
             k = coef(fit_time)[[1]], ols = summary(fit_time)$coef[1, 2],
             hac = nw_se(fit_time)[[1]], R2 = summary(fit_time)$r.squared),
  data.table(Bars = sprintf("%s-SOL volume bars", comma(VOL_BAR_SIZE)), n = nrow(bars_volume),
             k = coef(fit_volume)[[1]], ols = summary(fit_volume)$coef[1, 2],
             hac = nw_se(fit_volume)[[1]], R2 = summary(fit_volume)$r.squared)
))
fit_summary[, t_hac := k / hac]

knitr::kable(fit_summary, digits = c(0, 0, 8, 8, 8, 3, 1),
             format.args = list(big.mark = ","),
             col.names = c("Bars", "n", "$\\hat k$", "OLS s.e.", "HAC s.e.", "$R^2$", "$t$ (HAC)"),
             caption = "Permanent impact coefficient under both bar constructions.")
```

The HAC correction widens the standard error by about
`r sprintf("%.0f%%", 100 * (fit_summary$hac[1] / fit_summary$ols[1] - 1))` on time
bars. It does not change any conclusion — $t$ is still above 30 — but it is the
honest figure for overlapping bar data.

```{r}
#| label: fig-fits
#| fig-cap: "Bar-level mid-price change against net signed volume, under both bar constructions. The line is the fitted no-intercept regression."
#| code-summary: "Plot both fits"

scatter_data <- rbind(
  bars_time[,   .(imbalance, price_diff, Bars = sprintf("%d-minute time bars", BAR_MINUTES))],
  bars_volume[, .(imbalance, price_diff, Bars = sprintf("%s-SOL volume bars", comma(VOL_BAR_SIZE)))]
)
fit_lines <- data.table(
  Bars  = c(sprintf("%d-minute time bars", BAR_MINUTES),
            sprintf("%s-SOL volume bars", comma(VOL_BAR_SIZE))),
  slope = c(coef(fit_time)[[1]], coef(fit_volume)[[1]])
)

ggplot(scatter_data, aes(x = imbalance, y = price_diff)) +
  geom_point(alpha = 0.12, size = 0.7, colour = col_data) +
  geom_abline(data = fit_lines, aes(slope = slope, intercept = 0),
              colour = col_model, linewidth = 0.9) +
  facet_wrap(~ Bars, scales = "free_x") +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Mid-price change against order-flow imbalance",
    subtitle = sprintf("%s, %s — line is the fitted slope, k", SYMBOL, MONTH),
    x = "Imbalance (net signed SOL)", y = "Mid-price change (USD)"
  )
```

# Does the impact actually persist?

The model *assumes* the price move is permanent, but nothing so far tests it. If
the measured move were largely temporary — the book being pushed aside and then
refilling — it would decay once the bar ends, and $k$ would be an overstatement.

The test is direct: keep the same regressor, and push the point at which the
price is measured further into the future. `price_diff` is replaced by
$S_{t+h} - S_t$ for increasing $h$, with $S_t$ the bar's arrival mid. A genuinely
permanent impact leaves the slope flat; a decaying one does not.

```{r}
#| label: decay
#| code-summary: "Impact decay across measurement horizons"

horizons <- c(0, 1, 2, 3, 6, 12)   # in bars

decay <- rbindlist(lapply(horizons, function(h) {
  b <- copy(bars_time_all)
  b[, target := if (h == 0) price_diff else shift(pre_mid_price, -h) - pre_mid_price]
  f <- apply_filters(b)[!is.na(target)]
  m <- lm(target ~ imbalance + 0, f)
  data.table(h_bars = h, minutes = h * BAR_MINUTES,
             k = coef(m)[[1]], se = nw_se(m)[[1]],
             R2 = summary(m)$r.squared, n = nrow(f))
}))

knitr::kable(
  decay[, .(Window = fifelse(h_bars == 0, "bar end (as fitted above)",
                             sprintf("+%d minutes", minutes)),
            n, k, `HAC s.e.` = se, R2)],
  digits = c(0, 0, 8, 8, 4), format.args = list(big.mark = ","),
  col.names = c("Price measured at", "n", "$\\hat k$", "HAC s.e.", "$R^2$"),
  caption = "The estimate of k as the measurement point moves further past the bar.")
```

```{r}
#| label: fig-decay
#| fig-cap: "The impact coefficient as the price is measured further after the bar. The band is one HAC standard error. A flat profile means the price move does not revert."
#| fig-height: 3.8
#| code-summary: "Plot the decay profile"

ggplot(decay, aes(x = minutes, y = k)) +
  geom_hline(yintercept = coef(fit_time)[[1]], linetype = "dotted", colour = "grey50") +
  geom_ribbon(aes(ymin = k - se, ymax = k + se), fill = col_data, alpha = 0.18) +
  geom_line(colour = col_data, linewidth = 0.8) +
  geom_point(colour = col_data, size = 2.2) +
  expand_limits(y = 0) +
  scale_x_continuous(breaks = decay$minutes) +
  labs(
    title    = "The price move does not revert",
    subtitle = "k re-estimated with the price measured h minutes after the bar's arrival mid; dotted line = headline estimate",
    x = "Minutes after bar arrival", y = "Estimated k (USD per SOL)"
  )
```

**The impact is permanent.** From the bar end out to
`r max(decay[minutes <= 30]$minutes)` minutes, $\hat k$ moves from
`r format_sci(decay$k[1])` to `r format_sci(decay[minutes == 30]$k)` — a change
of `r sprintf("%.1f%%", 100 * (decay[minutes == 30]$k / decay$k[1] - 1))`, far
inside one standard error. Nothing reverts.

$R^2$ collapses over the same range (from
`r sprintf("%.3f", decay$R2[1])` to `r sprintf("%.3f", decay[minutes == 30]$R2)`)
but that is expected and is not evidence against permanence: as the horizon
grows, unrelated price movement accumulates in the dependent variable while the
signal stays the same size. The point estimate, not the fit quality, is what
carries the answer.

The mild rise by `r max(decay$minutes)` minutes is the well-known persistence of
order flow — imbalance is autocorrelated, so the later window captures the impact
of *subsequent* correlated flow as well, not extra impact from this bar.

# Is impact linear in imbalance?

The model asserts $dS = \sigma dW + k\,v\,dt$, which is linear in order flow.
That can be checked the same way the exponent was checked in the companion
notebook: group bars into equally populated imbalance bins and plot the
conditional mean.

```{r}
#| label: linearity
#| code-summary: "Binned conditional mean and power-law fit"

bin_imbalance <- function(b, n_bins = N_BINS) {
  d <- copy(b)[, ib := cut(imbalance,
                           breaks = unique(quantile(imbalance, seq(0, 1, length.out = n_bins + 1))),
                           include.lowest = TRUE)]
  d[!is.na(ib), .(mean_diff = mean(price_diff), mean_imb = mean(imbalance), n = .N),
    by = ib][order(mean_imb)]
}
binned_time <- bin_imbalance(bars_time)

# Power law, fitted in levels so that both signs are used: k |v|^beta sign(v).
power_time <- nls(price_diff ~ a * sign(imbalance) * abs(imbalance)^beta,
                  data = bars_time, start = list(a = 1e-3, beta = 0.7),
                  control = nls.control(maxiter = 300, warnOnly = TRUE))
power_volume <- nls(price_diff ~ a * sign(imbalance) * abs(imbalance)^beta,
                    data = bars_volume, start = list(a = 1e-3, beta = 0.7),
                    control = nls.control(maxiter = 300, warnOnly = TRUE))

beta_time <- coef(power_time)[["beta"]]
beta_vol  <- coef(power_volume)[["beta"]]
```

```{r}
#| label: fig-linearity
#| fig-cap: "Conditional mean price change by imbalance bin, against the linear and power-law fits. The data bend away from the straight line at both ends."
#| code-summary: "Plot linear vs power law"

grid <- data.table(imbalance = seq(min(bars_time$imbalance), max(bars_time$imbalance),
                                   length.out = 400))
grid[, Linear := coef(fit_time)[[1]] * imbalance]
grid[, `Power law` := coef(power_time)[["a"]] * sign(imbalance) * abs(imbalance)^beta_time]

ggplot() +
  geom_line(data = melt(grid, id.vars = "imbalance", variable.name = "Fit", value.name = "y"),
            aes(x = imbalance, y = y, colour = Fit, linetype = Fit), linewidth = 0.9) +
  geom_point(data = binned_time, aes(x = mean_imb, y = mean_diff),
             colour = col_data, size = 2.2) +
  scale_colour_manual(values = c("Linear" = col_muted, "Power law" = col_model)) +
  scale_linetype_manual(values = c("Linear" = "solid", "Power law" = "dashed")) +
  scale_x_continuous(labels = comma) +
  labs(
    title    = "Permanent impact is concave, not linear",
    subtitle = sprintf("%d imbalance bins of %s bars each; fitted exponent = %.2f",
                       nrow(binned_time), comma(round(mean(binned_time$n))), beta_time),
    x = "Imbalance (net signed SOL, bin mean)", y = "Mean mid-price change (USD)",
    caption = "Blue points: conditional mean per bin. The linear fit is the no-intercept regression above."
  )
```

```{r}
#| label: spec-table
#| code-summary: "Linear against power law"

train <- bars_time[isoweek(bin) <= 33]
test  <- bars_time[isoweek(bin) >  33]
m_lin <- lm(price_diff ~ imbalance + 0, train)
m_pow <- nls(price_diff ~ a * sign(imbalance) * abs(imbalance)^beta, train,
             start = list(a = 1e-3, beta = 0.7),
             control = nls.control(maxiter = 300, warnOnly = TRUE))
rmse <- function(e) sqrt(mean(e^2))

knitr::kable(
  data.table(
    # \lvert ... \rvert rather than |...|: a literal pipe inside a table cell is
    # a column separator to Pandoc, which escapes it and breaks the LaTeX math.
    Specification = c("Linear  $k\\,v$",
                      "Power law  $a\\,\\mathrm{sign}(v)\\lvert v\\rvert^{\\beta}$"),
    `Exponent`    = c(1, beta_time),
    AIC           = c(AIC(fit_time), AIC(power_time)),
    `Out-of-sample RMSE` = c(rmse(test$price_diff - predict(m_lin, test)),
                             rmse(test$price_diff - predict(m_pow, test)))
  ),
  digits = c(0, 3, 1, 4),
  caption = "Time bars. Out-of-sample fits ISO weeks 31-33 and evaluates on weeks 34-35.")
```

The exponent is `r sprintf("%.2f", beta_time)` (standard error
`r sprintf("%.3f", summary(power_time)$coef[2, 2])`) on time bars and
`r sprintf("%.2f", beta_vol)` on volume bars — both far below 1, and consistent
across the two bar constructions. The power law improves AIC by
`r sprintf("%.0f", AIC(fit_time) - AIC(power_time))` points and wins
out-of-sample, so it is not simply a more flexible model overfitting.

This does not mean $k$ should be discarded. The Almgren--Chriss solution needs
*linear* permanent impact for its closed form, and $\hat k$ remains the right
input to that framework. What the concavity means is that $\hat k$ is a **local
linearisation**, valid over the range of imbalances it was fitted on
(`r comma(round(min(bars_time$imbalance)))` to
`r comma(round(max(bars_time$imbalance)))` SOL per bar). Extrapolated beyond
that, it will overstate the impact of large flow.

# Filter sensitivity

The concavity has a direct consequence that is worth making explicit, because it
is the largest source of uncertainty in $\hat k$ — larger than the standard
error.

```{r}
#| label: sensitivity
#| code-summary: "Sensitivity to the cleaning choices"

spec_at <- function(b, label) {
  m <- lm(price_diff ~ imbalance + 0, b)
  p <- nls(price_diff ~ a * sign(imbalance) * abs(imbalance)^beta, b,
           start = list(a = 1e-3, beta = 0.7),
           control = nls.control(maxiter = 300, warnOnly = TRUE))
  data.table(Filter = label, n = nrow(b), k = coef(m)[[1]],
             beta = coef(p)[["beta"]], R2 = summary(m)$r.squared)
}

cut_sd  <- quantile(bars_time_all$std_dev,   TRIM_Q,     na.rm = TRUE)
cut_qty <- quantile(bars_time_all$base_qty,  TRIM_Q,     na.rm = TRUE)
cut_hi  <- quantile(bars_time_all$imbalance, TRIM_Q,     na.rm = TRUE)
cut_lo  <- quantile(bars_time_all$imbalance, 1 - TRIM_Q, na.rm = TRUE)

sens <- rbindlist(list(
  spec_at(bars_time_all[!is.na(std_dev)], "No trimming"),
  spec_at(bars_time_all[!is.na(std_dev) & std_dev <= cut_sd & base_qty <= cut_qty],
          "Volatility + volume only"),
  spec_at(bars_time_all[!is.na(std_dev) & imbalance <= cut_hi & imbalance >= cut_lo],
          "Imbalance tails only"),
  spec_at(bars_time, "All three (headline)"),
  spec_at(apply_filters(bars_time_all, 0.90), "All three, trimmed at 10/90")
))

knitr::kable(sens, digits = c(0, 0, 8, 3, 3), format.args = list(big.mark = ","),
             col.names = c("Filter", "n", "$\\hat k$", "$\\hat\\beta$", "$R^2$"),
             caption = "Both parameters under alternative cleaning choices.")
```

```{r}
#| label: fig-sensitivity
#| fig-cap: "The same five cleaning choices, expressed as a percentage deviation from the headline specification. Separate panels because the two parameters are on different scales — never a shared axis."
#| fig-height: 3.8
#| code-summary: "Plot the sensitivity"

base_k <- sens[Filter == "All three (headline)", k]
base_b <- sens[Filter == "All three (headline)", beta]

sens_long <- rbind(
  sens[, .(Filter, Parameter = "k (linear slope)",  dev = 100 * (k / base_k - 1))],
  sens[, .(Filter, Parameter = "beta (power-law exponent)", dev = 100 * (beta / base_b - 1))]
)
sens_long[, Filter := factor(Filter, levels = rev(sens$Filter))]
sens_long[, Parameter := factor(Parameter, levels = c("k (linear slope)",
                                                      "beta (power-law exponent)"))]

ggplot(sens_long, aes(x = dev, y = Filter, fill = Parameter)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.4) +
  geom_col(width = 0.6) +
  facet_wrap(~ Parameter) +
  scale_fill_manual(values = c("k (linear slope)" = col_data,
                               "beta (power-law exponent)" = col_model),
                    guide = "none") +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  labs(
    title    = "The exponent is stable where the slope is not",
    subtitle = "Deviation from the headline specification, in per cent",
    x = NULL, y = NULL
  )
```

Across these five defensible cleaning choices $\hat k$ moves by
`r sprintf("%.0f%%", 100 * (max(sens$k) / min(sens$k) - 1))`, while
$\hat\beta$ moves by only
`r sprintf("%.0f%%", 100 * (max(sens$beta) / min(sens$beta) - 1))`.

The mechanism is the concavity. The imbalance trim removes observations of the
*regressor*, and it removes them from the tails — which is exactly where a
concave curve is flattest. Cutting them off leaves a steeper average slope, so
$\hat k$ rises: `r format_sci(sens[Filter == "No trimming", k])` untrimmed
against `r format_sci(sens[Filter == "All three (headline)", k])` at the headline
setting. Under the power-law specification, which does not need the curve to be
straight, that sensitivity largely disappears. Any $\hat k$ quoted from this data
should therefore be quoted together with the trimming rule that produced it.

# Robustness

```{r}
#| label: robustness
#| code-summary: "Subsample and bar-size robustness"

weekly <- rbindlist(lapply(sort(unique(isoweek(bars_time$bin))), function(w) {
  b <- bars_time[isoweek(bin) == w]
  m <- lm(price_diff ~ imbalance + 0, b)
  data.table(Specification = sprintf("ISO week %d", w), n = nrow(b),
             k = coef(m)[[1]], R2 = summary(m)$r.squared)
}))

by_size <- rbindlist(lapply(c(2000, 5000, 20000), function(sz) {
  b <- apply_filters(make_bars(floor(trades$cum_base_qty / sz)))
  m <- lm(price_diff ~ imbalance + 0, b)
  data.table(Specification = sprintf("Volume bars, %s SOL", comma(sz)), n = nrow(b),
             k = coef(m)[[1]], R2 = summary(m)$r.squared)
}))

knitr::kable(rbind(weekly, by_size), digits = c(0, 0, 8, 3),
             format.args = list(big.mark = ","),
             col.names = c("Specification", "n", "$\\hat k$", "$R^2$"),
             caption = "Stability of k across subperiods and bar sizes.")
```

Week to week $\hat k$ ranges from `r format_sci(min(weekly$k))` to
`r format_sci(max(weekly$k))`, and across volume-bar sizes from
`r format_sci(min(by_size$k))` to `r format_sci(max(by_size$k))`. Both bands are
comparable to the spread produced by the cleaning choices, and both are much
wider than the HAC standard error — which is the usual situation for this kind of
estimate, and the reason the sensitivity table above matters more than the
regression output.

# Conclusion

The estimated permanent impact coefficient $k$ is
`r format_sci(coef(fit_time)[[1]])` for the time-bar specification and
`r format_sci(coef(fit_volume)[[1]])` for the volume-bar specification, in USD
per SOL of net signed flow. The two agree to within
`r sprintf("%.0f%%", 100 * abs(coef(fit_volume)[[1]] / coef(fit_time)[[1]] - 1))`,
so the estimate is not an artifact of the bar construction. In interpretable
terms, a net buy of 1,000 SOL — roughly
`r sprintf("$%s", comma(round(1000 * mean_price, -3)))` of notional — moves the
mid-price permanently by about
`r sprintf("%.2f", 1e4 * coef(fit_time)[[1]] * 1000 / mean_price)` bps.

Three findings qualify that number.

**The impact really is permanent.** Pushing the measurement point out to 30
minutes leaves $\hat k$ unchanged to within
`r sprintf("%.0f%%", abs(100 * (decay[minutes == 30]$k / decay$k[1] - 1)))`. This
is the assumption the whole model rests on, and it holds.

**But it is not linear.** The conditional mean is clearly concave, with an
exponent of `r sprintf("%.2f", beta_time)` on time bars and
`r sprintf("%.2f", beta_vol)` on volume bars. $\hat k$ should be read as a local
linearisation over the fitted imbalance range
(`r comma(round(min(bars_time$imbalance)))` to
`r comma(round(max(bars_time$imbalance)))` SOL per bar), appropriate as an
Almgren--Chriss input at those sizes, and increasingly conservative above them.

**The cleaning rule matters more than the standard error.** $\hat k$ varies by
`r sprintf("%.0f%%", 100 * (max(sens$k) / min(sens$k) - 1))` across reasonable
trimming choices against a HAC standard error of
`r sprintf("%.1f%%", 100 * nw_se(fit_time)[[1]] / coef(fit_time)[[1]])` of the
estimate. This is a consequence of forcing a straight line through a concave
relationship, and it is the reason the trimming rule belongs in any quotation of
$k$.

A caveat on external validity: this is one month of one instrument, and August
2024 contains the 5 August volatility event. The weekly breakdown is reassuring,
but confirming $\hat k$ and $\hat\beta$ on other months — `data/` holds January
through October — is the natural next step.
