---
title: "Estimation of Market Impact"
format: html
author: "Raj Kundi"
toc: TRUE
date: "08/21/2026"
---
```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false


library(ggplot2)
library(tidyverse)
library(data.table)
library(lubridate)
library(zoo)

# Format a number as "$m \times 10^{e}$" for correctly rendered scientific notation inline.
format_sci <- function(x, digits = 3) {
  exponent <- floor(log10(abs(x)))
  mantissa <- signif(x / 10^exponent, digits)
  sprintf("$%s \\times 10^{%d}$", format(mantissa), exponent)
}
```

# Introduction
To compute trading curves and trading envelopes, market impact and execution costs are needed. We estimate the permanent impact parameter using one month (August 2024) of Binance SOLUSD_PERP perpetual futures data: tick-level `trades` (individual executions, signed by taker side) and `bookTicker` (best bid/ask snapshots), from which the mid-price process is reconstructed.
```{r}
trades <- fread("data/SOLUSD_PERP-trades-2024-08.csv")
bookTicker <- fread("data/SOLUSD_PERP-bookTicker-2024-08.csv")
trades$ts <- as.POSIXct(
  trades$time / 1000,
  origin = "1970-01-01",
  tz = "UTC"
)
# Calculate mid-price
bookTicker[, mid_price := (best_bid_price + best_ask_price) / 2]
# Order the timestamps
setorder(trades, time)
setorder(bookTicker, transaction_time)
```
For each trade we attach the most recent mid-price observed before it (`pre_mid`) and the first mid-price observed after it (`post_mid`), so that the difference between the two isolates the price move associated with that trade (or a bar of trades) net of unrelated mid-price drift.
```{r}
setkey(bookTicker, transaction_time)
trades[, pre_mid := bookTicker[trades, on = .(transaction_time < time), mult = "last", .(mid_price)]$mid_price]
trades[, post_mid := bookTicker[trades, on = .(transaction_time > time), mult = "first", .(mid_price)]$mid_price]
```
The mid price of a stock is modeled by a process
$$
d S_t = \sigma d W_t + k v_t dt
$$
where $\sigma$ is a positive constant called the arithmetic volatility of the asset, and $k$ is the nonegative parameter modeling the magnitude of the **permanent market impact**.

Integrating the drift term over a bar gives $S_{t+\Delta} - S_t \approx k \sum_i v_i + \text{noise}$, so regressing a bar's mid-price change (`price_diff`) on its net signed volume (`imbalance`) recovers $k$, the permanent impact per unit of signed order flow, as the slope.

Bars are constructed two ways to check that the estimate of $k$ is not an artifact of the sampling scheme: fixed 5-minute time bars, and volume bars that close every 5,000 units of `base_qty` traded.
```{r}
trades[, sign := ifelse(is_buyer_maker, -1, 1)]
trades[, sign_base_qty := base_qty * sign]
bars_time <- trades[, .(
  imbalance = sum(sign_base_qty),
  base_qty = sum(base_qty),
  n_trades = .N,
  price_diff = last(post_mid) - first(pre_mid),
  pre_mid_price = first(pre_mid),
  post_mid_price = last(post_mid)
), by = .(bin = floor_date(ts, "5 minutes"))]
bars_time[, std_dev := rollapply(pre_mid_price, width = 15, FUN = sd, fill = NA, align = "right")]
```
```{r}
# Volume bars: group trades into buckets of 5000 base_qty traded
trades[, cum_base_qty := cumsum(base_qty)]
trades[, vol_bin := floor(cum_base_qty / 5000)]
bars_volume <- trades[, .(
  imbalance = sum(sign_base_qty),
  base_qty = sum(base_qty),
  n_trades = .N,
  price_diff = last(post_mid) - first(pre_mid),
  pre_mid_price = first(pre_mid),
  post_mid_price = last(post_mid)
), by = .(bin = vol_bin)]
bars_volume[, std_dev := rollapply(pre_mid_price, width = 15, FUN = sd, fill = NA, align = "right")]
```
The top 5% of rolling volatility (`std_dev`) and top 5% of traded volume (`base_qty`) are dropped to exclude bars dominated by news or liquidation events, the top and bottom 5% of `imbalance` are dropped to limit outlier influence on the OLS fit, and bars with `abs(imbalance) <= 1` (roughly below $100 notional) are dropped since they carry little order-flow information.
```{r}
filter_bars <- function(bars) {
  q_std_dev <- quantile(bars$std_dev, 0.95, na.rm = TRUE)
  q_base_qty <- quantile(bars$base_qty, 0.95, na.rm = TRUE)
  q_imbalance_high <- quantile(bars$imbalance, 0.95, na.rm = TRUE)
  q_imbalance_low <- quantile(bars$imbalance, 0.05, na.rm = TRUE)
  bars[
    std_dev <= q_std_dev &
      base_qty <= q_base_qty &
      imbalance <= q_imbalance_high &
      imbalance >= q_imbalance_low &
      abs(imbalance) > 1
  ]
}
bars_time <- filter_bars(bars_time)
bars_volume <- filter_bars(bars_volume)
```

# Regression
The slope on `imbalance` estimates $k$; its significance and the $R^2$ indicate how much of the bar-level mid-price change is explained by signed order flow alone.

## Time Bars (5 minutes)
```{r}
fitted_time <- lm(price_diff ~ imbalance + 0, bars_time)
summary(fitted_time)
```
```{r}
ggplot(bars_time, aes(x = imbalance, y = price_diff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Price Diff vs Imbalance (Time Bars)",
       x = "Imbalance", y = "Price Diff") +
  theme_minimal()
```

## Volume Bars (5000 base_qty)
```{r}
fitted_volume <- lm(price_diff ~ imbalance + 0, bars_volume)
summary(fitted_volume)
```
```{r}
ggplot(bars_volume, aes(x = imbalance, y = price_diff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Price Diff vs Imbalance (Volume Bars)",
       x = "Imbalance", y = "Price Diff") +
  theme_minimal()
```

# Conclusion
The estimated permanent impact coefficient $k$ (the slope on `imbalance`) is `r format_sci(coef(fitted_time)[["imbalance"]])` for the time-bar specification and `r format_sci(coef(fitted_volume)[["imbalance"]])` for the volume-bar specification. A stable, significant slope on `imbalance` across both the time-bar and volume-bar specifications would indicate a genuine permanent impact coefficient $k$ rather than an artifact of the chosen bar construction.