---
title: "Estimation of Market Impact"
format: html
author: "Raj Kundi"
toc: TRUE
date: "08/21/2026"
---
```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false
library(ggplot2)
library(tidyverse)
library(data.table)
library(lubridate)
library(zoo)
```

# Introduction
To compute trading curves and trading envelopes, market impact and execution costs are needed. 
```{r}
trades <- fread("data/SOLUSD_PERP-trades-2024-08.csv")
bookTicker <- fread("data/SOLUSD_PERP-bookTicker-2024-08.csv")
trades$ts <- as.POSIXct(
  trades$time / 1000,
  origin = "1970-01-01",
  tz = "UTC"
)
# Calculate mid-price
bookTicker[, mid_price := (best_bid_price + best_ask_price) / 2]
# Order the timestamps
setorder(trades, time)
setorder(bookTicker, transaction_time)
```
```{r}
setkey(bookTicker, transaction_time)
trades[, pre_mid := bookTicker[trades, on = .(transaction_time < time), mult = "last", .(mid_price)]$mid_price]
trades[, post_mid := bookTicker[trades, on = .(transaction_time > time), mult = "first", .(mid_price)]$mid_price]
```
The mid price of a stock is modeled by a process
$$
d S_t = \sigma d W_t + k v_t dt
$$
where $\sigma$ is a positive constant called the arithmetic volatility of the asset, and $k$ is the nonegative parameter modeling the magnitude of the permanent market impact. 

> investigate how the trader's own actions together with the order flow of other market participants affects the price of the assets she is trading.

```{r}
trades[, sign := ifelse(is_buyer_maker, -1, 1)]
trades[, sign_base_qty := base_qty * sign]
bars <- trades[, .(
  imbalance = sum(sign_base_qty),
  base_qty = sum(base_qty),
  n_trades = .N,
  price_diff = last(post_mid) - first(pre_mid),
  pre_mid_price = first(pre_mid),
  post_mid_price = last(post_mid)
), by = .(bin = floor_date(ts, "5 minutes"))]
bars[, std_dev := rollapply(pre_mid_price, width = 15, FUN = sd, fill = NA, align = "right")]
```
```{r}
q_std_dev <- quantile(bars$std_dev, 0.995, na.rm = TRUE)
q_imbalance_high <- quantile(bars$imbalance, 0.995, na.rm = TRUE)
q_imbalance_low <- quantile(bars$imbalance, 0.005, na.rm = TRUE)
bars <- bars[std_dev <= q_std_dev & imbalance <= q_imbalance_high & imbalance >= q_imbalance_low]
```

# Regression
```{r}
fitted <- lm(price_diff ~ imbalance, bars)
summary(fitted)
```
```{r}
ggplot(bars, aes(x = bin, y = std_dev)) +
  geom_line() +
  labs(title = "Rolling Standard Deviation", x = "Time", y = "Std Dev")
```
```{r}
ggplot(bars, aes(x = imbalance, y = price_diff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Price Diff vs Imbalance",
       x = "Imbalance", y = "Price Diff")
```