---
title: "Market Impact on ETHUSDT Spot Market"
format: html
author: "Raj Kundi"
date: "08/20/2026"
toc: true
---

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

# load libraries
library("dplyr")
library("tidyr")
library("lubridate")
library("tibble")
library("knitr")
library("ggplot2")
library(data.table)

# Load data
trades <- read.csv("../data/ETHUSDT-trades-2026-07.csv")
trades$ts <- as_datetime(trades$time/1000)
trades <- arrange(trades, id)
```


# Estimating market impact

R. Almgren, C. Thum, E. Hauptmann, and H. Li. Direct estimation of equity market impact. Risk, 18(7):58-62, 2005

## Introduction

Transaction costs have a big impact on investment performance (i.e. Freyre-Sanders, Guobuzaite and Byrne (2004)). It also impacts unwinding positions. Costs generally fall into two categories:

- *Direct costs* are commissions and fees that are explicitly stated and easily measured. 
- *Indirect costs* are costs that are not explicitly stated. For large trades, the most important component of these is the impact of a trader's own actions on the market.

## Data

```{r}
library(data.table)
library(lubridate)

setDT(trades)
setorder(trades, ts)

# --- 1. Resample onto a regular 5-minute grid ---
freq <- "5 mins"
bars_per_day <- 24 * 12   # 288 five-minute bars per day

grid <- data.table(
  ts = seq(
    floor_date(min(trades$ts), "5 mins"),
    ceiling_date(max(trades$ts), "5 mins"),
    by = freq
  )
)

setkey(trades, ts)
setkey(grid, ts)

# rolling join: last known trade price at/before each grid timestamp
px <- trades[grid, roll = TRUE, .(ts, price)]
px <- px[!is.na(price)]

# --- 2. 24h return at each grid point ---
px[, ret_24h := log(price / shift(price, bars_per_day))]

# --- 3. Rolling 2-week std dev of those overlapping 24h returns ---
window_bars <- 14 * bars_per_day   # 4032 five-minute bars = 14 days

px[, vol_24h := frollapply(ret_24h, n = window_bars, FUN = sd, na.rm = TRUE, align = "right")]

# annualise
px[, vol_annualised := vol_24h * sqrt(365)]

# --- 4. Join return and volatility back onto each trade ---
vol_lookup <- px[!is.na(vol_24h), .(ts, ret_24h, vol_24h, vol_annualised)]
setkey(vol_lookup, ts)
setkey(trades, ts)

trades[, c("ret_24h", "vol_24h", "vol_annualised") :=
         vol_lookup[trades, .(ret_24h, vol_24h, vol_annualised), roll = TRUE]]
```

Drop trades during high variance or very small order sizes. 

### Variables

The goal is to descrive market impact in terms of a small number of input variables. Here 