---
title: "Market Impact on BTCUSDT Futures Market"
format: html
author: "Raj Kundi"
date: "08/20/2026"
toc: true
---

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

library("dplyr")
library("tidyr")
library("lubridate")
library("tibble")
library("knitr")
library("ggplot2")
```

# BTCUSDT perpetual market

## Binance historical data

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

trades <- read.csv("../data/BTCUSDT-trades-2026-08-16.csv")
trades$ts <- as_datetime(trades$time/1000)
trades <- arrange(trades, id)
bookdepth <- read.csv("../data/BTCUSDT-bookdepth-2026-08-16.csv")
metrics <- read.csv("../data/BTCUSDT-metrics-2026-08-16.csv")
```


| Attribute/field | data |
|----|----|
| `id` | Unique identifier for each trade. |
| `price` | Price the trade happened at. |
| `qty`, `base_qty` | Quantity traded. |
| `time` | Timestamp the trade happened at. |
| `is_buyer_maker` | Was the buyer the passive trader, or was a bid limit order lifted. |


## Generate orderflow imbalance measurements and estimate market impact parameters

We take buy or sell from the prespective of the aggresive trader placing market orders.

```{r}
#| echo: false
#| paged-print: true

generate_imbalance_bars <- function(data, timescale) {
  data |> mutate(
  bar = floor_date(ts, "5 minutes"),
  buy_qty = if_else(is_buyer_maker == "true", qty, 0),
  sell_qty = if_else(is_buyer_maker == "false", qty, 0),
  buy_quote_qty = if_else(is_buyer_maker == "true", quote_qty, 0),
  sell_quote_qty = if_else(is_buyer_maker == "false", quote_qty, 0),
  price = price
) |>
  group_by(bar) |>
  summarise(
    n_trades = n(),
    buy_qty = sum(buy_qty),
    buy_quote_qty = sum(buy_quote_qty),
    sell_qty = sum(sell_qty),
    sell_quote_qty = sum(sell_quote_qty),
    open = first(price),
    high = max(price),
    low = min(price),
    close = last(price),
    .groups = "drop"
  ) |>
  mutate(
    qty = buy_qty + sell_qty,
    quote_qty = buy_quote_qty + sell_quote_qty,
    sell_proportion = sell_qty / qty,
    imbalance = abs(buy_qty - sell_qty) / qty,
    price_change = (close - open),
  )
}

# 1. Generate orderflow imbalance measurements
imbalance_bars <- generate_imbalance_bars(trades, "5 minutes") |>
  subset(price_change > 0 & imbalance > 0)


# 2. Fit the log-log linear model
fit <- lm(log(abs(price_change)) ~ log(abs(imbalance)), data = imbalance_bars)

# 3. Extract parameter estimates and back-transform the intercept
a_param <- exp(coef(fit)[1])
b_param <- coef(fit)[2]

# 4. Extract 95% confidence intervals and back-transform the intercept bounds
ci <- confint(fit, level = 0.95)
lower_a <- exp(ci["(Intercept)", "2.5 %"])
upper_a <- exp(ci["(Intercept)", "97.5 %"])
lower_b <- ci["log(abs(imbalance))", "2.5 %"]
upper_b <- ci["log(abs(imbalance))", "97.5 %"]

# 5. Create a smooth continuous grid over your x-axis to build the envelope
x_grid <- seq(min(abs(imbalance_bars$imbalance)), 
              max(abs(imbalance_bars$imbalance)), 
              length.out = 500)

# 6. Construct the envelope curves dataframe
envelope_df <- data.frame(
  imbalance = x_grid,
  fitted_curve = a_param * (x_grid ^ b_param),
  lower_bound = lower_a * (x_grid ^ lower_b),
  upper_bound = upper_a * (x_grid ^ upper_b)
)

# 7. Render the final market impact visualization
ggplot() +
  # A. Raw Data Points
  geom_point(data = imbalance_bars,
             aes(x = abs(imbalance), y = abs(price_change)),
             alpha = 0.4, color = "darkblue") +
  
  # B. Translucent Parameter Envelope (Confidence Ribbon)
  geom_ribbon(data = envelope_df, 
              aes(x = imbalance, ymin = lower_bound, ymax = upper_bound), 
              fill = "red", alpha = 0.15) +
  
  # C. Central Fitted Power Curve Estimate
  geom_line(data = envelope_df, 
            aes(x = imbalance, y = fitted_curve), 
            color = "red", linewidth = 1.2) +
  
  # D. Clarifying Labels & Clean Theme
  labs(
    title = "Market Impact: Imbalance vs Price Change (95% confidence interval)",
    subtitle = paste0("Fitted Power Curve: y = ", round(a_param, 2), " * x^", round(b_param, 4)),
    x = "Absolute Imbalance",
    y = "Absolute Price Change"
  ) +
  theme_minimal()
```
