---
title: "Estimation of the Execution Cost Function"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 8
    fig-height: 4.5
author: "Raj Kundi"
date: "08/22/2026"
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

library(ggplot2)
library(tidyverse)
library(data.table)
library(lubridate)
library(zoo)
library(scales)

setDTthreads(0)

# Format a number as "$m \times 10^{e}$" for correctly rendered scientific notation inline.
format_sci <- function(x, digits = 3) {
  exponent <- floor(log10(abs(x)))
  mantissa <- signif(x / 10^exponent, digits)
  sprintf("$%s \\times 10^{%d}$", format(mantissa), exponent)
}

# ---- Plot styling -----------------------------------------------------------
# Categorical hues are slots 1-3 of a colourblind-safe reference palette, used in
# fixed order: blue = measured data, orange = fitted model, grey = context.
col_data  <- "#2a78d6"
col_model <- "#eb6834"
col_muted <- "#b4b3ac"

theme_set(theme_minimal(base_size = 11))
theme_update(
  panel.grid.minor = element_blank(),
  panel.grid.major = element_line(linewidth = 0.3, colour = "grey88"),
  plot.title       = element_text(face = "bold", size = 12),
  plot.subtitle    = element_text(colour = "grey35", size = 10),
  plot.caption     = element_text(colour = "grey45", size = 8, hjust = 0),
  strip.text       = element_text(colour = "grey25", size = 9),
  legend.position  = "bottom",
  legend.title     = element_blank()
)
```

```{r}
#| label: parameters
#| code-summary: "Estimation parameters"

SYMBOL      <- "SOLUSD_PERP"
MONTH       <- "2024-08"   # estimation sample
WARMUP      <- "2024-07"   # read for daily volume only, to warm up the trailing ADV

BAR_MINUTES <- 5     # bar length over which the participation rate is defined
ADV_WINDOW  <- 10    # trailing days in the average-daily-volume estimate of V_t
VOL_WINDOW  <- 15    # bars in the rolling volatility estimate used for filtering
TRIM_Q      <- 0.95  # drop the top 5% of rolling volatility and of bar volume
N_BINS      <- 20    # participation-rate quantile bins for the conditional mean
```

# Introduction

In the Almgren--Chriss framework the market volume process is $V_t$, and a trader
executing at rate $v_t$ pays, per base asset transacted at time $t$,

$$
S_t + g\!\left(\frac{v_t}{V_t}\right),
$$

where $S_t$ is the mid-price. If the trader buys the trading price is above the
mid-price; if she sells it is below. Writing $\rho = v_t / V_t$ for the
**participation rate**, the price paid per token is

$$
S_t + g(\rho) = S_t + \eta\,\operatorname{sign}(\rho)\,|\rho|^{\phi} + \psi\,\operatorname{sign}(\rho).
$$

Rather than working with $g$ directly it is convenient to use the **execution
cost function** $L(\rho) = \rho\, g(\rho)$, which measures cost per unit time
rather than per unit traded. If $X_t$ is the trader's cash account process, then

$$
\begin{array}{rcl}
d X_t & = & - v_t \left( S_t + g\!\left( \frac{v_t}{V_t} \right) \right) dt \\[4pt]
      & = & -v_t S_t \, dt - V_t\, L\!\left( \frac{v_t}{V_t} \right) dt .
\end{array}
$$

$L$ is taken to be a strictly convex power function,

$$
L(\rho) = \eta\,|\rho|^{1 + \phi} + \psi\,|\rho|, \qquad \eta, \psi > 0,\ \phi > 0,
$$

whose two terms are economically distinct.

**The proportional term $\psi|\rho|$** is linear, and is the per-token add-on: a
constant mark-up or mark-down paid on every asset traded, no matter how fast or
slow the trader goes. It is a *fixed transaction cost* — the effective spread,
the venue fee, and any taxes or similar charges — not a liquidity cost.

**The super-linear term $\eta|\rho|^{1+\phi}$** models **temporary market
impact**. When a trader trades faster she consumes more liquidity and the price
moves against her own trading pressure. This is the term that makes the optimal
execution problem non-trivial, and estimating its two parameters is the main work
of this notebook.

This notebook estimates $\psi$, $\eta$ and $\phi$ from one month of Binance
`r SYMBOL` perpetual futures data. It is the companion to
`../market_impact/estimation_of_market_impact.qmd`, which estimates the
*permanent* impact coefficient $k$ from the same data; together the two
parameterise the trading-curve problem.

## What each parameter is measured against

The two cost terms are separated by choosing the right price benchmark, and
getting this right is what makes the estimation work at all:

| Parameter | Measured from | Measured to | Interpretation |
|---|---|---|---|
| $\psi$ | mid-price | best bid/ask | half the quoted spread: the cost of crossing, paid even by an infinitesimally small order |
| $\eta,\ \phi$ | best bid/ask | realised price | **slippage**: the cost of walking the book beyond the touch |

Measuring slippage from the *touch* rather than from the mid is what stops the
two terms absorbing one another. An order that fills entirely at the best quote
has zero slippage by construction and contributes only $\psi$.

# Data

## Sources

Two tick-level Binance feeds for `r SYMBOL`, `r MONTH`:

- **`trades`** — individual executions. Each row is a *fill*, carrying `price`,
  `base_qty` (size in SOL), millisecond `time`, and `is_buyer_maker`, which signs
  the taker side: `is_buyer_maker == TRUE` means the *taker was selling* into the
  bid; `FALSE` means the taker was buying from the ask.
- **`bookTicker`** — best bid/ask snapshots, from which the mid-price process and
  the spread are reconstructed. Only the three columns needed are read, since the
  raw file is several gigabytes.

```{r}
#| label: load-data
#| message: false
#| code-summary: "Load trades and bookTicker"

trades <- fread(sprintf("data/%s-trades-%s.csv", SYMBOL, MONTH))
bookTicker <- fread(
  sprintf("data/%s-bookTicker-%s.csv", SYMBOL, MONTH),
  select = c("best_bid_price", "best_ask_price", "transaction_time")
)

trades[, ts   := as.POSIXct(time / 1000, origin = "1970-01-01", tz = "UTC")]
trades[, date := as.Date(ts)]
# is_buyer_maker == TRUE  =>  the maker was the buyer  =>  the TAKER sold.
trades[, taker_buy := !is_buyer_maker]

bookTicker[, mid_price     := (best_bid_price + best_ask_price) / 2]
bookTicker[, quoted_spread := best_ask_price - best_bid_price]
# Half the quoted spread: what a marketable order concedes just by crossing.
bookTicker[, half_spread   := quoted_spread / 2]
bookTicker[, ts := as.POSIXct(transaction_time / 1000, origin = "1970-01-01", tz = "UTC")]

setorder(trades, time, id)
setorder(bookTicker, transaction_time)
```

The sample contains `r comma(nrow(trades))` fills and
`r comma(nrow(bookTicker))` quote updates spanning
`r format(min(trades$ts), "%Y-%m-%d")` to `r format(max(trades$ts), "%Y-%m-%d")`.

## Attaching the prevailing quote

Each fill is matched to the last quote strictly *before* it. The strict
inequality matters: a quote stamped at the same millisecond as the fill may
already reflect that fill's own consumption of the book, which would mechanically
shrink the measured slippage toward zero.

```{r}
#| label: quote-join
#| code-summary: "Non-equi join of fills to the prevailing quote"

setkey(bookTicker, transaction_time)

trades[, pre_bid := bookTicker[trades, on = .(transaction_time < time),
                               mult = "last", .(best_bid_price)]$best_bid_price]
trades[, pre_ask := bookTicker[trades, on = .(transaction_time < time),
                               mult = "last", .(best_ask_price)]$best_ask_price]
trades[, pre_mid := (pre_bid + pre_ask) / 2]
```

```{r}
#| label: fig-spread
#| fig-cap: "Half the quoted spread through the month. The instrument sits at the minimum tick almost all of the time; the visible excursions are the news and liquidation episodes that the filtering step below is designed to remove."
#| code-summary: "Plot the spread process"

spread_plot <- bookTicker[seq(1, .N, by = 2000)]
spread_plot[, half_spread_ma := rollmean(half_spread, k = 200, fill = NA, align = "right")]

ggplot(spread_plot, aes(x = ts)) +
  geom_line(aes(y = half_spread, colour = "Raw"), alpha = 0.5, linewidth = 0.25) +
  geom_line(aes(y = half_spread_ma, colour = "Rolling mean"), linewidth = 0.6, na.rm = TRUE) +
  scale_colour_manual(values = c("Raw" = col_muted, "Rolling mean" = col_data)) +
  # Log scale: the series sits at the minimum tick for most of the month, so a
  # linear axis is entirely taken up by a handful of spikes.
  scale_y_log10(labels = label_number(drop0trailing = TRUE)) +
  annotation_logticks(sides = "l", colour = "grey70", linewidth = 0.25) +
  labs(
    title    = "Half the quoted spread",
    subtitle = sprintf("%s, %s — every 2,000th quote update, rolling mean over 200 sampled points",
                       SYMBOL, MONTH),
    x = NULL, y = "Half spread (USD per SOL)"
  )
```

```{r}
#| label: free-bookticker
#| include: false

# The quote series is no longer needed once the fills carry their arrival quotes.
rm(bookTicker, spread_plot)
invisible(gc())

# Share of timestamps carrying both taker directions: justifies grouping orders
# by timestamp AND side rather than by timestamp alone.
pct_mixed <- trades[, .(n_sides = uniqueN(taker_buy)), by = time][, 100 * mean(n_sides > 1)]
```

# Reconstructing marketable orders

A single marketable order that sweeps several price levels is reported by the
exchange as **several fill rows sharing one millisecond timestamp**. Treating
those rows as separate trades destroys precisely the signal we are after: the
first fragment executes at the touch with zero slippage, and only the later
fragments show the concession. Fills are therefore aggregated into orders, and
the order's realised price is the size-weighted average of its fills,

$$
\bar{P} = \frac{\sum_i q_i p_i}{\sum_i q_i}, \qquad
\text{slippage} =
\begin{cases}
\bar{P} - \text{ask}^{-} & \text{taker buys,} \\[2pt]
\text{bid}^{-} - \bar{P} & \text{taker sells,}
\end{cases}
$$

where $\text{ask}^{-}$ and $\text{bid}^{-}$ are the quotes prevailing before the
order's first fill. Grouping is by timestamp *and* taker side rather than by
timestamp alone: only `r sprintf("%.2f%%", pct_mixed)` of timestamps contain both
directions, but merging a buy with a sell would net opposing flow into a
meaningless average price.

```{r}
#| label: build-orders
#| code-summary: "Aggregate fills into marketable orders"

trades[, order_id := rleid(time, taker_buy)]

orders <- trades[, .(
  ts        = first(ts),
  date      = first(date),
  taker_buy = first(taker_buy),
  base_qty  = sum(base_qty),
  notional  = sum(base_qty * price),
  n_fills   = .N,
  pre_bid   = first(pre_bid),
  pre_ask   = first(pre_ask),
  pre_mid   = first(pre_mid)
), by = order_id]

orders[, exec_price  := notional / base_qty]
orders[, slippage    := fifelse(taker_buy, exec_price - pre_ask, pre_bid - exec_price)]
orders[, half_spread := (pre_ask - pre_bid) / 2]
setorder(orders, ts)
```

```{r}
#| label: order-stats
#| include: false

n_fills_tot <- nrow(trades)
n_orders    <- nrow(orders)
pct_zero    <- 100 * mean(orders$slippage == 0)
pct_neg     <- 100 * mean(orders$slippage <  0)
pct_pos     <- 100 * mean(orders$slippage >  0)

# Counterfactual used in the "Bar construction" section: if a bar's cost were
# measured as its VWAP against the bar's OPENING mid, the accumulated price
# drift over the bar would swamp the impact signal and turn the cost negative
# this often.
pct_drift_neg <- orders[, .(vwap = sum(notional) / sum(base_qty),
                            mid0 = first(pre_mid), tb = first(taker_buy)),
                        by = .(bin = floor_date(ts, sprintf("%d minutes", BAR_MINUTES)),
                               taker_buy)
][, 100 * mean(fifelse(tb, vwap - mid0, mid0 - vwap) < 0)]
```

`r comma(n_fills_tot)` fills collapse to `r comma(n_orders)` orders. The
resulting slippage distribution is dominated by a point mass at exactly zero.

```{r}
#| label: fig-slippage-dist
#| fig-cap: "Order slippage relative to the touch. The bar at zero is not bad data — it is every order that filled entirely at the best quote, and it holds most of the sample."
#| fig-height: 3.6
#| code-summary: "Plot the order slippage distribution"

dist <- orders[, .(bucket = fcase(
  slippage <  0, "Negative",
  slippage == 0, "Exactly zero",
  default        = "Positive"
))][, .N, by = bucket]
dist[, share  := N / sum(N)]
dist[, bucket := factor(bucket, levels = c("Negative", "Exactly zero", "Positive"))]

ggplot(dist, aes(x = bucket, y = share)) +
  geom_col(fill = col_data, width = 0.55) +
  geom_text(aes(label = percent(share, accuracy = 0.1)),
            vjust = -0.6, size = 3.4, colour = "grey25") +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.14))) +
  labs(
    title    = "Most marketable orders never leave the touch",
    subtitle = sprintf("%s orders, %s", comma(n_orders), MONTH),
    x = NULL, y = "Share of orders"
  )
```

`r sprintf("%.1f%%", pct_zero)` of orders execute entirely at the best quote and
so have exactly zero slippage, `r sprintf("%.1f%%", pct_pos)` pay a positive
concession, and `r sprintf("%.1f%%", pct_neg)` are negative. The negative values
are a measurement artifact rather than a free lunch: they arise when the book
improves between the quote snapshot and the fill, or when hidden liquidity rests
inside the spread.

::: {.callout-note}
## Why the zeros are kept here but dropped later

Zero-slippage orders carry no information about the *shape* of $g$, and in a
log-log regression they cannot even be represented. But they are genuine
zero-cost executions, and the quantity being modelled is the *average* cost of
executing a given quantity. Discarding them before averaging would overstate the
cost of every bar that contains one.

They are therefore retained inside the volume-weighted bar average and dropped
only at the bar level, where they are rare. The alternative — dropping them first
— is estimated and reported in the robustness table.
:::

# The proportional cost $\psi$

$\psi$ is the per-token cost that does not depend on trading speed. It has two
components:

1. **The effective spread** — the distance from the mid-price to the touch, which
   is half the quoted spread. This is directly measurable and is the part
   estimated here.
2. **The venue fee.** Fees are *not* embedded in the `price` field of the trades
   feed, so there is nothing in this data to estimate. The fee is a known constant
   for any given account tier and simply adds to $\psi$.

The spread is size-weighted, since what matters for execution cost is the spread
prevailing when volume actually trades, not the spread averaged over idle time.

```{r}
#| label: psi
#| code-summary: "Estimate the proportional cost"

psi_vw     <- orders[, sum(base_qty * half_spread) / sum(base_qty)]
psi_median <- median(orders$half_spread)
mean_price <- orders[, sum(notional) / sum(base_qty)]

psi_vw_bps     <- 1e4 * psi_vw / mean_price
psi_median_bps <- 1e4 * psi_median / mean_price
```

| Estimate | USD per SOL | bps of mid |
|---|---|---|
| Volume-weighted mean half-spread | `r sprintf("%.6f", psi_vw)` | `r sprintf("%.3f", psi_vw_bps)` |
| Median half-spread | `r sprintf("%.6f", psi_median)` | `r sprintf("%.3f", psi_median_bps)` |

The volume-weighted figure is an order of magnitude above the median because size
arrives when the spread is wide. Since $\psi$ is meant to price the cost actually
paid, the volume-weighted estimate

$$
\hat\psi = `r sprintf("%.6f", psi_vw)` \ \text{USD per SOL} \approx `r sprintf("%.2f", psi_vw_bps)` \ \text{bps}
$$

is the one carried forward. It excludes fees, which on a typical taker tier are
one to two orders of magnitude larger than the spread on this instrument.

# The participation rate

## Why the unit of analysis must be a bar

$\rho = v_t / V_t$ is a ratio of two *rates*: the trader's execution rate and the
market's volume rate, over the same horizon. A single order's size divided by a
whole day's volume is not that ratio, and it produces $\rho \approx 10^{-6}$ —
five orders of magnitude below the participation rates any real execution
schedule operates at. The section [Why not estimate at the order level?] shows
empirically that the exponent is simply not identified at that scale.

The estimation is therefore run on **`r BAR_MINUTES`-minute bars, split by taker
side**. Within a bar, one side's takers are treated as a single trader executing a
schedule: $v$ is that side's total volume and $V$ is the market's volume rate over
the same window. This mirrors the bar construction used for permanent impact in
the companion notebook.

## The market volume rate $V_t$

$V_t$ is a trailing `r ADV_WINDOW`-day average daily volume, lagged one day so
that no part of the estimate uses information from the day it is applied to. The
preceding month is read for its daily volumes only, so the trailing window is
fully populated from the first day of the sample.

```{r}
#| label: adv
#| message: false
#| code-summary: "Trailing average daily volume"

warmup <- fread(sprintf("data/%s-trades-%s.csv", SYMBOL, WARMUP),
                select = c("base_qty", "time"))
warmup[, date := as.Date(as.POSIXct(as.numeric(time) / 1000,
                                    origin = "1970-01-01", tz = "UTC"))]

daily_volume <- rbind(
  warmup[, .(volume = sum(base_qty)), by = date],
  trades[,  .(volume = sum(base_qty)), by = date]
)[, .(volume = sum(volume)), by = date][order(date)]

# Trailing mean, then lagged one day: strictly causal.
daily_volume[, adv := shift(rollapply(volume, ADV_WINDOW, mean,
                                      fill = NA, align = "right"), 1)]

rm(warmup)
invisible(gc())
```

```{r}
#| label: fig-adv
#| fig-cap: "Daily traded volume against the lagged trailing average that defines the market volume rate. The 5 August spike is the yen-carry unwind; the trailing average deliberately absorbs it slowly."
#| fig-height: 3.6
#| code-summary: "Plot daily volume against the trailing ADV"

ggplot(daily_volume[date >= as.Date(paste0(MONTH, "-01"))], aes(x = date)) +
  geom_col(aes(y = volume, fill = "Daily volume"), width = 0.7) +
  geom_line(aes(y = adv, colour = "Trailing ADV (lagged)"), linewidth = 0.8, na.rm = TRUE) +
  scale_fill_manual(values   = c("Daily volume" = col_muted)) +
  scale_colour_manual(values = c("Trailing ADV (lagged)" = col_data)) +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  labs(
    title    = "Daily volume and the market volume rate",
    subtitle = sprintf("%s, %s — %d-day trailing mean, lagged one day",
                       SYMBOL, MONTH, ADV_WINDOW),
    x = NULL, y = "Volume (SOL)"
  )
```

## Bar construction

For each bar and taker side,

$$
\rho = \frac{\sum_j q_j}{\mathrm{ADV} \times \Delta / 1440},
\qquad
\text{slippage}_{\text{bar}} = \frac{\sum_j q_j\, s_j}{\sum_j q_j},
$$

with $q_j$ and $s_j$ the size and slippage of order $j$ and $\Delta$ the bar
length in minutes. The denominator of $\rho$ scales the daily volume rate down to
the bar, so $\rho = 1$ means that side alone traded at the market's own average
rate.

Each order's slippage is measured against *its own* arrival quote, so the bar
average is free of the price drift that accumulates over the bar. That drift is
permanent impact, and belongs to the companion notebook rather than here. The
distinction is not academic: measuring each bar's VWAP against the bar's
*opening* mid instead confounds the two, and the resulting cost comes out
**negative in `r sprintf("%.0f%%", pct_drift_neg)` of bars** — the drift over the
bar simply swamps the impact signal.

```{r}
#| label: make-bars
#| code-summary: "Bar construction and filtering functions"

make_bars <- function(ord, minutes, drop_zero_orders = FALSE) {
  d <- if (drop_zero_orders) ord[slippage != 0] else ord
  d <- copy(d)[, bin := floor_date(ts, sprintf("%d minutes", minutes))]

  # Arrival mid of each bar, and the rolling volatility used for filtering.
  arrival <- d[, .(bar_mid = first(pre_mid)), by = bin][order(bin)]
  arrival[, sd_mid := rollapply(bar_mid, VOL_WINDOW, sd, fill = NA, align = "right")]

  b <- d[, .(
    date     = first(date),
    qty      = sum(base_qty),
    n_orders = .N,
    slip_vw  = sum(base_qty * slippage) / sum(base_qty)
  ), by = .(bin, taker_buy)]

  b <- merge(merge(b, arrival, by = "bin"), daily_volume[, .(date, adv)], by = "date")
  b[, rho := qty / (adv * (minutes / (24 * 60)))]
  b[]
}

apply_filters <- function(b, q = TRIM_Q) {
  cut_sd  <- quantile(b$sd_mid, q, na.rm = TRUE)
  cut_qty <- quantile(b$qty,    q, na.rm = TRUE)
  b[!is.na(sd_mid) & sd_mid <= cut_sd & qty <= cut_qty & rho > 0 & slip_vw > 0]
}

bars_all <- make_bars(orders, BAR_MINUTES)
bars     <- apply_filters(bars_all)
```

# Data cleaning

Three filters are applied, following the same logic as the permanent-impact
notebook.

**Top `r sprintf("%.0f%%", 100 * (1 - TRIM_Q))` of rolling volatility.** When the
mid-price is moving violently the book at the touch is thin and reprices between
the quote snapshot and the fill. Measured slippage in those bars is dominated by
that repricing rather than by the trader's own consumption of liquidity, so with
respect to $g$ it is noise. Volatility is the rolling standard deviation of the
bar arrival mid over `r VOL_WINDOW` bars.

**Top `r sprintf("%.0f%%", 100 * (1 - TRIM_Q))` of bar volume.** The heaviest bars
are liquidation cascades and news reactions. They are genuine observations, but
they sit so far out in $\rho$ that they would dominate the fit, and their cost
dynamics are not the ones a normal execution schedule faces.

**Non-positive bar slippage.** A bar whose volume-weighted slippage is zero or
negative cannot enter a log-log regression. At bar level this is rare, because
averaging washes out the individual measurement artifacts, so the truncation bias
is small. That is emphatically not true at order level, where the same rule would
discard `r sprintf("%.0f%%", pct_zero + pct_neg)` of the sample.

```{r}
#| label: fig-filters
#| fig-cap: "The two trimming thresholds. Bars to the right of each dashed line are dropped."
#| fig-height: 3.6
#| code-summary: "Plot the filter thresholds"

cut_sd  <- quantile(bars_all$sd_mid, TRIM_Q, na.rm = TRUE)
cut_qty <- quantile(bars_all$qty,    TRIM_Q, na.rm = TRUE)

metric_levels <- c("Rolling volatility of the mid (USD)", "Bar volume (SOL)")

filt_long <- rbind(
  bars_all[!is.na(sd_mid) & sd_mid > 0, .(value = sd_mid, metric = metric_levels[1])],
  bars_all[qty > 0,                     .(value = qty,    metric = metric_levels[2])]
)
filt_long[, metric := factor(metric, levels = metric_levels)]

cuts <- data.table(metric = factor(metric_levels, levels = metric_levels),
                   cut    = c(cut_sd, cut_qty))

ggplot(filt_long, aes(x = value)) +
  geom_histogram(bins = 70, fill = col_data, alpha = 0.85) +
  geom_vline(data = cuts, aes(xintercept = cut),
             colour = col_model, linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~ metric, scales = "free") +
  scale_x_log10(labels = label_number(drop0trailing = TRUE, big.mark = ",")) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = sprintf("Trimming the top %.0f%% of volatility and of volume",
                       100 * (1 - TRIM_Q)),
    subtitle = sprintf("%d-minute bars, log x-axis. Dashed line = %.0fth percentile",
                       BAR_MINUTES, 100 * TRIM_Q),
    x = NULL, y = "Bars"
  )
```

```{r}
#| label: cleaning-table
#| code-summary: "Cleaning waterfall"

n_drop_sd   <- bars_all[is.na(sd_mid) | sd_mid > cut_sd, .N]
n_drop_qty  <- bars_all[!is.na(sd_mid) & sd_mid <= cut_sd & qty > cut_qty, .N]
n_drop_slip <- bars_all[!is.na(sd_mid) & sd_mid <= cut_sd & qty <= cut_qty & slip_vw <= 0, .N]

knitr::kable(
  data.table(
    Step = c("Bars constructed",
             sprintf("Drop top %.0f%% rolling volatility", 100 * (1 - TRIM_Q)),
             sprintf("Drop top %.0f%% bar volume",         100 * (1 - TRIM_Q)),
             "Drop non-positive slippage",
             "Retained"),
    Bars = c(nrow(bars_all), -n_drop_sd, -n_drop_qty, -n_drop_slip, nrow(bars))
  ),
  format.args = list(big.mark = ","),
  caption = "Effect of each cleaning step on the bar sample."
)
```

`r sprintf("%.1f%%", 100 * nrow(bars) / nrow(bars_all))` of bars survive. The
retained sample spans participation rates from `r sprintf("%.3f", min(bars$rho))`
to `r sprintf("%.2f", max(bars$rho))`, with a median of
`r sprintf("%.2f", median(bars$rho))` — a range that covers the schedules an
execution algorithm would actually run.

# Estimating $\eta$ and $\phi$

Taking logs of $g(\rho) = \eta \rho^{\phi}$ linearises the power law, so the slope
of $\log(\text{slippage})$ on $\log \rho$ estimates $\phi$ and the intercept
estimates $\log \eta$.

```{r}
#| label: fit
#| code-summary: "Fit the power law"

fit <- lm(log(slip_vw) ~ log(rho), data = bars)
summary(fit)

eta <- exp(coef(fit)[[1]])
phi <- coef(fit)[[2]]
```

```{r}
#| label: fig-scatter
#| fig-cap: "Bar-level slippage against participation rate on log-log axes. A pure power law is a straight line in these coordinates."
#| code-summary: "Plot slippage against participation rate"

# A few dozen near-zero bars stretch the axis over five decades and squash the
# bulk into a sliver, so the view is zoomed to the central 99.8%. This is a zoom
# only -- coord_cartesian clips the view, it does not drop rows, and the fit
# above was estimated on every retained bar.
y_zoom  <- quantile(bars$slip_vw, c(0.001, 0.999))
n_off   <- bars[slip_vw < y_zoom[1] | slip_vw > y_zoom[2], .N]

ggplot(bars, aes(x = rho, y = slip_vw)) +
  geom_point(alpha = 0.12, size = 0.7, colour = col_data) +
  geom_abline(intercept = log10(eta), slope = phi,
              colour = col_model, linewidth = 0.9) +
  scale_x_log10(labels = label_number(accuracy = 0.01)) +
  scale_y_log10(labels = label_number(drop0trailing = TRUE)) +
  annotation_logticks(sides = "bl", colour = "grey70", linewidth = 0.25) +
  coord_cartesian(ylim = y_zoom) +
  labs(
    title    = "Slippage against participation rate",
    subtitle = sprintf("%d-minute bars by taker side, n = %s; fitted exponent = %.3f",
                       BAR_MINUTES, comma(nrow(bars)), phi),
    x = expression(rho == v[t]/V[t]),
    y = "Volume-weighted slippage (USD per SOL)",
    caption = sprintf(paste("Orange line: the fitted power law. View zoomed to the central 99.8%% of bars;",
                            "%d bars fall outside it and are included in the fit."), n_off)
  )
```

The scatter is wide, as it must be — a single bar's realised cost is one draw from
a heavy-tailed distribution. The systematic relationship is far clearer in the
**conditional mean**, obtained by grouping bars into `r N_BINS` equally populated
participation-rate bins and averaging within each. Averaging is the right
operation: it is the conditional mean of $g$ that the model describes, and it
cancels the symmetric measurement noise that the log transform otherwise
distorts.

```{r}
#| label: binned
#| code-summary: "Binned conditional mean"

bars[, rho_bin := cut(rho,
                      breaks = unique(quantile(rho, seq(0, 1, length.out = N_BINS + 1))),
                      include.lowest = TRUE)]

binned <- bars[!is.na(rho_bin), .(
  mean_slippage   = mean(slip_vw),
  median_slippage = median(slip_vw),
  rho_med         = median(rho),
  n               = .N
), by = rho_bin][order(rho_med)]

fit_binned <- lm(log(mean_slippage) ~ log(rho_med), data = binned, weights = binned$n)
eta_binned <- exp(coef(fit_binned)[[1]])
phi_binned <- coef(fit_binned)[[2]]
```

```{r}
#| label: fig-binned
#| fig-cap: "Conditional mean slippage by participation-rate bin, against the fitted power law. Each bin holds an equal number of bars."
#| code-summary: "Plot the binned conditional mean"

grid <- data.table(rho = 10^seq(log10(min(binned$rho_med)),
                                log10(max(binned$rho_med)), length.out = 200))
grid[, pred := eta_binned * rho^phi_binned]

ggplot() +
  geom_line(data = grid, aes(x = rho, y = pred, colour = "Fitted power law"),
            linewidth = 0.9, linetype = "dashed") +
  geom_point(data = binned, aes(x = rho_med, y = mean_slippage, colour = "Conditional mean"),
             size = 2.2) +
  scale_colour_manual(values = c("Conditional mean" = col_data,
                                 "Fitted power law" = col_model)) +
  scale_x_log10(labels = label_number(accuracy = 0.01)) +
  scale_y_log10(labels = label_number(accuracy = 0.0001)) +
  annotation_logticks(sides = "bl", colour = "grey70", linewidth = 0.25) +
  labs(
    title    = "The conditional mean is a clean power law",
    subtitle = sprintf("%d bins of %s bars each; exponent = %.3f, R-squared = %.3f",
                       nrow(binned), comma(round(mean(binned$n))),
                       phi_binned, summary(fit_binned)$r.squared),
    x = expression(rho~"(bin median)"),
    y = "Mean slippage (USD per SOL)"
  )
```

The binned fit reaches $R^2 =$ `r sprintf("%.3f", summary(fit_binned)$r.squared)`,
against `r sprintf("%.3f", summary(fit)$r.squared)` for the unbinned regression.
The gap is not a contradiction: the power law describes the conditional mean well,
while individual bars scatter widely around it. Both estimates of $\phi$ are
reported.

# Robustness

```{r}
#| label: robustness
#| code-summary: "Robustness checks"

summarise_fit <- function(f, label) {
  m <- lm(log(slip_vw) ~ log(rho), f)
  data.table(Specification = label, n = nrow(f),
             eta = exp(coef(m)[[1]]), phi = coef(m)[[2]],
             R2 = summary(m)$r.squared)
}

rob <- rbindlist(list(
  # Sampling horizon
  rbindlist(lapply(c(1, 5, 15, 30, 60), function(mm)
    summarise_fit(apply_filters(make_bars(orders, mm)), sprintf("Bar length: %d min", mm)))),
  # Strength of the trimming
  rbindlist(lapply(c(1, 0.99, 0.95, 0.90), function(q)
    summarise_fit(apply_filters(bars_all, q), sprintf("Trim: keep %.0f%%", 100 * q)))),
  # Zero-slippage orders removed before aggregating
  summarise_fit(apply_filters(make_bars(orders, BAR_MINUTES, drop_zero_orders = TRUE)),
                "Zero-slippage orders dropped first"),
  # Subsamples
  rbindlist(lapply(sort(unique(isoweek(bars$bin))), function(w)
    summarise_fit(bars[isoweek(bin) == w], sprintf("ISO week %d", w)))),
  summarise_fit(bars[taker_buy == TRUE],  "Taker buys only"),
  summarise_fit(bars[taker_buy == FALSE], "Taker sells only")
))

knitr::kable(rob, digits = c(0, 0, 5, 3, 3),
             format.args = list(big.mark = ","),
             col.names = c("Specification", "n", "$\\hat\\eta$", "$\\hat\\phi$", "$R^2$"),
             caption = "Sensitivity of the estimates to sampling and cleaning choices.")
```

```{r}
#| label: fig-robust
#| fig-cap: "The exponent across bar lengths under both estimators. The conditional-mean estimator is stable throughout and drifts toward the square-root value as the horizon lengthens."
#| fig-height: 3.6
#| code-summary: "Plot exponent stability"

bl <- rbindlist(lapply(c(1, 5, 15, 30, 60), function(mm) {
  f  <- apply_filters(make_bars(orders, mm))
  m  <- lm(log(slip_vw) ~ log(rho), f)
  f2 <- copy(f)[, rb := cut(rho, unique(quantile(rho, seq(0, 1, length.out = N_BINS + 1))),
                            include.lowest = TRUE)]
  bn <- f2[!is.na(rb), .(ms = mean(slip_vw), r = median(rho), n = .N), by = rb]
  mb <- lm(log(ms) ~ log(r), bn, weights = bn$n)
  data.table(minutes = mm,
             `Unbinned regression`  = coef(m)[[2]],
             `Conditional-mean fit` = coef(mb)[[2]])
}))

ggplot(melt(bl, id.vars = "minutes", variable.name = "Estimator", value.name = "phi"),
       aes(x = minutes, y = phi, colour = Estimator)) +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 1, y = 0.5, label = "square-root law", vjust = -0.6,
           hjust = 0, size = 3, colour = "grey45") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = c("Unbinned regression"  = col_data,
                                 "Conditional-mean fit" = col_model)) +
  scale_x_log10(breaks = c(1, 5, 15, 30, 60)) +
  expand_limits(y = 0.4) +
  labs(
    title    = "Exponent stability across bar length",
    subtitle = "The conditional-mean estimator does not depend on the sampling horizon",
    x = "Bar length (minutes, log scale)", y = "Estimated exponent"
  )
```

The estimates hold up. Buys and sells agree to within
`r sprintf("%.3f", abs(rob[Specification == "Taker buys only", phi] - rob[Specification == "Taker sells only", phi]))`,
and tightening the trim from keeping 100% of bars to keeping 90% moves $\phi$ by
less than
`r sprintf("%.2f", abs(rob[Specification == "Trim: keep 100%", phi] - rob[Specification == "Trim: keep 90%", phi]))`.

Two entries deserve comment. The 1-minute unbinned fit is the one clear outlier,
at $\phi =$ `r sprintf("%.2f", rob[Specification == "Bar length: 1 min", phi])`; at
that horizon many bars contain only a handful of orders, so `slip_vw` is often
close to zero and the log transform amplifies the resulting noise. The
conditional-mean estimator at the same horizon gives
`r sprintf("%.2f", bl[minutes == 1][["Conditional-mean fit"]])`, in line with every
other bar length. Separately, dropping zero-slippage orders *before* aggregating
raises $\hat\eta$ from `r sprintf("%.5f", eta)` to
`r sprintf("%.5f", rob[Specification == "Zero-slippage orders dropped first", eta])`,
exactly as expected: removing the cost-free executions inflates every bar average.
It is reported for completeness but not used.

# Why not estimate at the order level?

It is natural to want $\rho$ defined per marketable order, and the reason that
does not work is worth recording. The obstacle is scale. The median order is
`r sprintf("%.1f", median(orders$base_qty))` SOL against an average daily volume of
roughly `r sprintf("%.1f", mean(daily_volume$adv, na.rm = TRUE) / 1e6)`M SOL, so
$\rho \approx 10^{-6}$. Essentially every order fills within one or two book
levels, and its slippage is pinned by the tick size rather than driven by its size
— there is no dynamic range for the exponent to be identified from.

```{r}
#| label: order-level
#| code-summary: "Order-level fit, for comparison"

orders[, adv       := daily_volume[match(orders$date, date)]$adv]
orders[, rho_order := base_qty / adv]

ord_sample <- orders[slippage > 0 & rho_order > 0]
ord_fit    <- lm(log(slippage) ~ log(rho_order), ord_sample)

ord_weeks <- rbindlist(lapply(sort(unique(isoweek(ord_sample$ts))), function(w) {
  m <- lm(log(slippage) ~ log(rho_order), ord_sample[isoweek(ts) == w])
  data.table(Week = w, phi = coef(m)[[2]], R2 = summary(m)$r.squared)
}))

knitr::kable(ord_weeks, digits = 4,
             col.names = c("ISO week", "$\\hat\\phi$", "$R^2$"),
             caption = sprintf(
               "Order-level exponent by week. Full-sample fit: phi = %.3f, R-squared = %.4f, n = %s.",
               coef(ord_fit)[[2]], summary(ord_fit)$r.squared, comma(nrow(ord_sample))))
```

The full-sample fit explains
`r sprintf("%.1f%%", 100 * summary(ord_fit)$r.squared)` of the variation, and the
weekly estimates range from `r sprintf("%.2f", min(ord_weeks$phi))` to
`r sprintf("%.2f", max(ord_weeks$phi))` — the parameter is not identified.
Aggregating to bars fixes this because it lifts $\rho$ into a range where size,
rather than the tick, determines cost.

# Conclusion

The fitted execution cost function for `r SYMBOL` over `r MONTH` is

$$
L(\rho) = \hat\eta\,|\rho|^{1+\hat\phi} + \hat\psi\,|\rho|,
\qquad
\hat\eta = `r sprintf("%.5f", eta)`,\quad
\hat\phi = `r sprintf("%.3f", phi)`,\quad
\hat\psi = `r sprintf("%.5f", psi_vw)`,
$$

with the conditional-mean estimator giving $\hat\phi =$
`r sprintf("%.3f", phi_binned)` as the more robust alternative. All figures are in
USD per SOL. $\hat\psi$ covers the spread only; the venue fee is a known constant
that adds to it.

```{r}
#| label: fig-cost
#| fig-cap: "The estimated per-token cost of crossing, decomposed. Below the crossover the spread dominates; above it, temporary impact does."
#| code-summary: "Plot the fitted cost function"

crossover <- (psi_vw / eta)^(1 / phi)

curve <- data.table(rho = 10^seq(log10(min(bars$rho)), log10(max(bars$rho)),
                                 length.out = 300))
curve[, `Temporary impact`  := 1e4 * eta * rho^phi / mean_price]
curve[, `Proportional cost` := 1e4 * psi_vw / mean_price]
curve[, `Total`             := `Temporary impact` + `Proportional cost`]

ggplot(melt(curve, id.vars = "rho", variable.name = "Component", value.name = "bps"),
       aes(x = rho, y = bps, colour = Component)) +
  geom_vline(xintercept = crossover, linetype = "dotted", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = crossover, y = max(curve$Total),
           label = sprintf("  crossover at rho = %.2f", crossover),
           hjust = 0, size = 3, colour = "grey35") +
  scale_colour_manual(values = c("Temporary impact"  = col_model,
                                 "Proportional cost" = col_muted,
                                 "Total"             = col_data)) +
  scale_x_log10(labels = label_number(accuracy = 0.01)) +
  labs(
    title    = "Estimated cost of crossing the spread, per token",
    subtitle = sprintf("%s, %s — eta = %.5f, phi = %.3f, psi = %.5f USD/SOL",
                       SYMBOL, MONTH, eta, phi, psi_vw),
    x = expression(rho == v[t]/V[t]), y = "Cost (bps of mid-price)"
  )
```

Three things are worth drawing out.

**The exponent is close to the square-root law.** $\hat\phi \approx$
`r sprintf("%.2f", phi_binned)`–`r sprintf("%.2f", phi)`, converging on $0.5$ as
the bar length grows. That $L(\rho) \propto \rho^{1+\phi}$ is strictly convex is
what the optimal-execution problem requires, and the condition holds comfortably:
$\hat\phi > 0$ with a standard error of
`r sprintf("%.3f", summary(fit)$coefficients[2, 2])`.

**Spread and impact cross over near $\rho =$ `r sprintf("%.2f", crossover)`.**
Below that participation rate the fixed cost of crossing dominates the cost of
walking the book, so a schedule's binding constraint is how *often* it crosses,
not how hard it pushes. Above it, impact dominates and the trade-off against
timing risk becomes the live one. For context, the median bar in this sample sits
at $\rho =$ `r sprintf("%.2f", median(bars$rho))`, below the crossover.

**Costs on this instrument are small in absolute terms.** At $\rho = 1$ — one side
trading at the market's entire average rate — the total per-token cost is about
`r sprintf("%.2f", 1e4 * (psi_vw + eta) / mean_price)` bps before fees. For any
schedule operating below the crossover, the venue fee is the larger part of the
total cost.

A caveat on external validity: this is one month of one instrument, and August
2024 contains the 5 August volatility event. The weekly breakdown in the
robustness table is reassuring, but confirming $\hat\phi$ on other months and
other instruments — both available in `data/` — is the natural next step.
